Less Funding, More Lecturers, and Fewer Professors: \\ \vspace{0.1cm} Stagnating State Funding for Higher Education and its Effect on Faculty

### 2024-03-15 text/sections/04-empirical.tex
updated

The conditional independence assumption for the shift-share instrument makes sense in theory for state funding, but is fundamentally untestable with data.
However, the assumption is bolstered by the fact that universities with higher values of the shift-share instrument differ little in observed characteristics compared to those with lower values \citep{pei2019poorly}.
Universities are by-in-large smaller (in terms of enrolment, total revenues, and professor count) at the top of the shift-share instrument distribution (see~\autoref{tab:summary-quantiles}).
Though on a per student basis, there is little difference across the distribution (except in the outcomes under consideration).
The shift-share instrument is positively associated with the total enrolment and total amount of state funding for each university, in both \$ and $\log$/percentage change terms, while other the other sources of university finances are not associated with the funding shift-share.
So that the other sources of finances are not clear confounders for the IV strategy, as they exhibit balance with respect to the funding shift-share \citep{pei2019poorly}.
Yearly and individual fixed effects are included in regressions throughout to implicitly condition on mean differences between universities and years.

\subsection{Instrumental Variables (IV) Model}
\label{sec:iv-model-uni}

I use the instrument defined in \autoref{eqn:public-instrument} to overcome the endogeneity concerns for state funding to each public university, for an IV approach.
\autoref{eqn:firststage} is the first stage for exogenous variation in the state funding for university $i$ in year $t$, and \autoref{eqn:secondstage} the second stage for the effect of state funding on faculty outcomes.\footnote{
    The IV approach identifies the local average treatment effect, weighted towards larger values of the instrument in the continuous case.
    So we interpret this treatment effect as a weighted average of causal effects on faculty at public universities to state funding changes, specific to shift-share instruments, among compliers --- i.e., among universities who respond to funding cut and would not have made faculty-outcome changes absent the funding cut.
    Also, we assume that no universities state funding increased in response to the negative shift-share instrument (monotonicity).
}
\begin{equation}
    \label{eqn:secondstage}
    Y_{i,t} = \alpha_i + \gamma_t + \beta \widehat X_{i,t} + \varepsilon_{i,t}
\end{equation}
I estimate the system by two stage least squares, including institution and year fixed effects.\footnote{
    Note that dividing by student count also implicitly controls for the size of the university, so that this model implicitly accounts for yearly variation in professor count and university revenues arising from a university's size of growth/decline.
}
$Y_{i,t}$ represents faculty outcomes for university $i$ in year $t$, $\alpha_i, \gamma_t$ university and year fixed effects, and $\widehat X_{i,t}$ state funding for university $i$ estimated in first stage \eqref{eqn:firststage}.

>>>>

The conditional independence assumption for the shift-share instrument makes sense in theory for state funding, but is fundamentally untestable with data.
However, the assumption is bolstered by the fact that universities with higher values of the shift-share instrument differ little in observed characteristics compared to those with lower values \citep{pei2019poorly}.
Universities are by-in-large smaller (in terms of enrolment, total revenues, and professor count) at the top of the shift-share instrument distribution (see~\autoref{tab:summary-quantiles}).
Though on a per student basis, there is little difference across the distribution (except in the outcomes under consideration).
The shift-share instrument is positively associated with the total enrolment and total amount of state funding for each university, in both \$ and $\log$/percentage change terms, while other the other sources of university finances are not associated with the funding shift-share.
So that the other sources of finances are not clear confounders for the IV strategy, as they exhibit balance with respect to the funding shift-share.
Nonetheless, yearly and institution fixed effects are included in regressions throughout to implicitly condition on mean differences between universities and years.

\subsection{Instrumental Variables (IV) Model}
\label{sec:iv-model-uni}
I use the instrument defined in \autoref{eqn:public-instrument} to overcome the endogeneity concerns for state funding to each public university, for an IV approach.
\autoref{eqn:firststage} is the first stage for exogenous variation in the state funding for university $i$ in year $t$, and \autoref{eqn:secondstage} the second stage for the effect of state funding on faculty outcomes.\footnote{
    The IV approach identifies the local average treatment effect, weighted towards larger values of the instrument in the continuous case.
    So we interpret this treatment effect as a weighted average of causal effects on faculty at public universities to state funding changes, specific to shift-share instruments, among compliers --- i.e., among universities who respond to funding cut and would not have made faculty-outcome changes absent the funding cut.
    Also, we assume that no universities state funding increased in response to the negative shift-share instrument (monotonicity).
}
\begin{equation}
    \label{eqn:secondstage}
    Y_{i,t} = \alpha_i + \gamma_t + \beta \widehat X_{i,t} + \varepsilon_{i,t}
\end{equation}
I estimate the system by two stage least squares, including institution and year fixed effects.\footnote{
    Note that dividing by student count also implicitly controls for the size of the university, so that this model implicitly accounts for yearly variation in professor count and university revenues arising from a university's size of growth/decline.
}
$Y_{i,t}$ represents faculty outcomes for university $i$ in year $t$, $\alpha_i, \gamma_t$ university and year fixed effects, and $\widehat X_{i,t}$ state funding for university $i$ estimated in first stage \eqref{eqn:firststage}.

### 2024-03-15 text/sections/05-results.tex
updated

\subsection{Reduced Hiring of New Professors}
The results above present a condundrum: there are large substitution effects among faculty, but faculty themselves are unaffected by funding cuts.
The analysis of Illinois faculty (IBHED) studies the effect of state funding cuts on faculty already at the university --- i.e., incumbent faculty.
The findings above imply that subsitution towards lecturers came about by limiting hiring of new professors, to replace leaving or retiring professors, so that they are replaced by hiring lecturers.

To further study this issue, I take recent data on the total faculty hired at US public universities over 2011--2021 provided by \cite{wapman2022quantifying,wapman2022zenodo}.
\cite{wapman2022quantifying} study the composition of professor hiring in the years 2011--2021, and publiccally released data on the number of professors (not lecturers) hired at US universities.\footnote{
    These data are not panel data, instead only give the total professors hired over the ten year period 2011--2021.
}
I combine this measure, with total state funding for public universities acorss the same time period from IPEDS, to validate the claim that state funding cuts led to faculty subtitution via limiting hiring of new professors.

>>>>

\subsection{Reduced Hiring of New Professors}
The results above present a conundrum: there are large substitution effects among faculty, but faculty themselves are unaffected by funding cuts.
The analysis of Illinois faculty (IBHED) studies the effect of state funding cuts on faculty already at the university --- i.e., incumbent faculty.
The findings above imply that substitution towards lecturers came about by limiting hiring of new professors, to replace leaving or retiring professors, so that they are replaced by hiring lecturers.
In this subsection, I test this implication with data on universities' hiring of new professors.

I take recent data on the total faculty hired at US public universities over 2011--2021 provided by \cite{wapman2022quantifying,wapman2022zenodo}.
\cite{wapman2022quantifying} study the composition of professor hiring in the years 2011--2021, and publicly released data on the number of professors (not lecturers) hired at US universities.\footnote{
    These data are not panel data, instead only give the total professors hired over the ten-year period 2011--2021.
}
I combine this measure, with total state funding for public universities across the same time period from IPEDS, to validate the claim that state funding cuts led to faculty substitution via limiting hiring of new professors.

### 2024-03-15 text/sections/06-discussion.tex
updated

%\textbf{implication: public unis are increasing sheer count of lecturers, in response to enrolment, and not their tenured faculty.
%SO prof per student is where the movemenet is most stark.}
Universities received less state funding, relative to increasing enrolment, and reacted by increasing their reliance on lecturers relative to assistant and full professors in both the short and medium-run; these results imply that lecturers can be considered substitutes for professors in the tenure-system.\footnote{
    Additional results in Appendix \autoref{sec:appendix-substitution} formalises the marginal rate of substitution between lecturers and professors on tenure-track, showing how assistant and full professors were substituted for lecturers while there is no substitution between assistant and full professors.
}
The average public university's state funding fell by 36\% per student, while the number of lecturers per student increased by 113\%, and assistant and full professors  fell by 17\% and 23\%, respectively.
Stagnating state funding explains about 40\% of the observed substitution towards lecturers away from tenure-track and tenured professors.

>>>>

%\textbf{implication: public unis are increasing sheer count of lecturers, in response to enrolment, and not their tenured faculty.
%SO prof per student is where the movemenet is most stark.}
Universities received less state funding, relative to increasing enrolment, and reacted by increasing their reliance on lecturers relative to assistant and full professors in both the short and medium-run; these results imply that lecturers can be considered substitutes for professors in the tenure-system.
%\footnote{
%    Additional results in Appendix \autoref{sec:appendix-substitution} formalises the marginal rate of substitution between lecturers and professors on tenure-track, showing how assistant and full professors were substituted for lecturers while there is no substitution between assistant and full professors.
%}
The average public university's state funding fell by 36\% per student, while the number of lecturers per student increased by 113\%, and assistant and full professors  fell by 17\% and 23\%, respectively.
Stagnating state funding explains about 40\% of the observed substitution towards lecturers away from tenure-track and tenured professors.

### 2024-03-15 text/sections/07-conclusion.tex
updated

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Conclusion section
\section{Summary and Concluding Remarks}
\label{sec:conclusion}

This paper studies how stagnating state funding for higher education affected faculty, and their composition, at US public universities.
This work contributes to the literature along two primary dimensions.
First, by isolating changes in state funding on public universities via state funding shocks, this work provides an explanation for the increased reliance on lecturers and away from tenure-track and tenured professors at US public universities.
This approach used multiple methods to estimate the short- and medium-run effects of state funding cuts on faculty outcomes.
Second, this work investigates how individual faculty are affected by changes in state funding for their university using a dataset new in the economic literature (IBHED).
These data allow for detailed analysis of thousands of faculty salaries, and employment outcomes in the Illinois university system.

Public universities have systematically substituted away from tenure-track and tenured professors, toward lecturers, in the face of persistent declines in state funding for higher education.
The effects on incumbent professors in Illinois are non-distinguishable from zero, with suggestive evidence for reduced hiring of new professors, implying that the faculty substitution is driven by reduced hiring of new professors.
Public universities are using more contingent lecturers to teach their students, while private universities continue to employ more tenure-track and tenured professors than their public counterparts, and each year the gap widens.

At the same time as higher education costs have been rising in the US, public universities have also dealt with declining state funding.
I show that these headwinds led to systematic change at public universities, changes that affect their faculty, and likely limiting public universities in their goals in research and education.
These results show large changes in faculty composition, and that stagnation in state support explains at least a third of the observed shift away from tenured professors and towards lecturers.
At the same time, private universities were not exposed to financial headwinds of the same magnitude or persistence.
While public universities continue to educate the majority of higher education students in the US, we should worry about the effects of restricting their funding has on faculty, and the wider impact on higher education as a whole.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Conclusion section
\section{Summary and Concluding Remarks}
\label{sec:conclusion}
Public universities have systematically substituted away from tenure-track and tenured professors, toward lecturers, in the face of persistent declines in state funding for higher education.
The effects on incumbent professors in Illinois are non-distinguishable from zero, and suggestive evidence shows that funding cuts limited hiring of new professors, implying that substitution towards lecturers is driven by reduced hiring of new professors.
Public universities are using more contingent lecturers to teach their students, while private universities continue to employ more tenure-track and tenured professors than their public counterparts, and each year the gap widens.

This work contributes to the literature along two primary dimensions.
First, this work provides an explanation for the increased reliance on lecturers and away from tenure-track and tenured professors at US public universities, by isolating the effects funding cuts from state-wide funding shifts in both the short-- and long-run.
Second, this work investigates how individual faculty are affected by changes in state funding for their university using a dataset new in the economic literature (IBHED).
These data allow for detailed analysis of thousands of faculty salaries, and employment outcomes, in the Illinois university system, and show that incumbent faculty were relatively unaffected by state funding cuts to their employers.

While state funding for higher education has stagnated, costs of education has been rising at the same time, so that funding cuts have even more impact today than they did three decades ago.
I show that these financial headwinds led to systematic change at public universities, leading to a substitution away from stable employment contracts (tenure, professorship) and towards less stable arrangements (adjunct contracts, lecture positions).
These results show large changes in faculty composition, and that stagnation in state support explains at least a third of the observed shift away from tenured professors and towards lecturers.
This likely had knock-on effects on instruction, contribiting to worsening student outcomes at US public universities since the early 1990s --- while private universities were not exposed to financial headwinds of the same magnitude or persistence over the same time period.
While public universities educate the majority of US higher education students, we should worry about the effects of restricting their funding has on faculty, and the wider impact on higher education as a whole.

### 2024-03-15 text/sections/10-appendix.tex
updated

\newpage
\subsection{Rates of Substitution}
\label{sec:appendix-substitution}

The funding elasticities can be used to recover the marginal rate of substitution between two outcomes.
For example, write $Y^1$ for the number of lecturers per student at a university, and $Y^2$ for the number of full professors.
I use the above approaches to estimate the funding elasticities, where $\% \Delta$ denotes percent change.
\[ \beta_1 = \frac{\% \Delta Y^1}{\% \Delta X}
\text{, and }
\beta_2 = \frac{\% \Delta Y^2}{\% \Delta X} \]
As such, it is possible to recover the elasticity for substitution between lecturers and full professors by the universities via the respective funding elasticities.
\[ \frac{\% \Delta Y^1}{\% \Delta Y^2}
= \frac{\% \Delta Y^1 / \% \Delta X}{\% \Delta Y^2 / \% \Delta X}
= \frac{\beta_1}{\beta_2} \]
I present results for the rates of substitution between different levels of faculty by this approach, dividing the relevant coefficient estimates and presenting standard errors calculated by a non-parametric bootstrap.
In practice, this corresponds to division of the estimates of the elasticity for employment of professors (by rank) with respect to state funding, presented in Panel B \autoref{tab:facultycount-shock-reg}, and bootstrapping the results to generate standard errors and confidence intervals.
% https://stackoverflow.com/questions/63777368/computing-the-standard-error-when-dividing-coefficients-of-different-regressions

The implied marginal rate of substitution between lecturers and assistant professors is estimated as -3.26 (standard error 0.50), based on 10,000 bootstrap samples.
This means that public universities increased their number lecturers per student by 3.26\% when they decreased their count of assistant professors, on average and subject to the changes in state funding they experienced 1990--2017.
Between lecturers and full professors the rate of substitution is -3.19 (0.34), which implies that universities substitute between lecturers and full professors in the same way.
Between assistant and full professors the rate of substitution is 0.99 (0.11), which intuitively implies that universities treated assistant and full professors (i.e., those before and after tenure in the tenure system) as complements.

%"Calculated point est for substitution between lecturers + ast profs" 
%                                                                      
%                                                  "-3.26372574229461" 
%                                                                      
%                                                           "with SEs" 
%                                                                      
%                                                  "0.501721750806993" 
%                                                                      
%                                                        "and 95 % CI" 
%                                                                 2.5% 
%                                                  "-4.38380704259523" 
%                                                                97.5% 
%                                                  "-2.40372605889863"

>>>>

%\newpage
%\subsection{Rates of Substitution}
%\label{sec:appendix-substitution}
%The funding elasticities can be used to recover the marginal rate of substitution between two outcomes.
%For example, write $Y^1$ for the number of lecturers per student at a university, and $Y^2$ for the number of full professors.
%I use the above approaches to estimate the funding elasticities, where $\% \Delta$ denotes percent change.
%\[ \beta_1 = \frac{\% \Delta Y^1}{\% \Delta X}
%\text{, and }
%\beta_2 = \frac{\% \Delta Y^2}{\% \Delta X} \]
%As such, it is possible to recover the elasticity for substitution between lecturers and full professors by the universities via the respective funding elasticities.
%\[ \frac{\% \Delta Y^1}{\% \Delta Y^2}
%= \frac{\% \Delta Y^1 / \% \Delta X}{\% \Delta Y^2 / \% \Delta X}
%= \frac{\beta_1}{\beta_2} \]
%I present results for the rates of substitution between different levels of faculty by this approach, dividing the relevant coefficient estimates and presenting standard errors calculated by a non-parametric bootstrap.
%In practice, this corresponds to division of the estimates of the elasticity for employment of professors (by rank) with respect to state funding, presented in Panel B \autoref{tab:facultycount-shock-reg}, and bootstrapping the results to generate standard errors and confidence intervals.
%% https://stackoverflow.com/questions/63777368/computing-the-standard-error-when-dividing-coefficients-of-different-regressions
%
%The implied marginal rate of substitution between lecturers and assistant professors is estimated as -3.26 (standard error 0.50), based on 10,000 bootstrap samples.
%This means that public universities increased their number lecturers per student by 3.26\% when they decreased their count of assistant professors, on average and subject to the changes in state funding they experienced 1990--2017.
%Between lecturers and full professors the rate of substitution is -3.19 (0.34), which implies that universities substitute between lecturers and full professors in the same way.
%Between assistant and full professors the rate of substitution is 0.99 (0.11), which intuitively implies that universities treated assistant and full professors (i.e., those before and after tenure in the tenure system) as complements.
%
%"Calculated point est for substitution between lecturers + ast profs" 
%                                                                      
%                                                  "-3.26372574229461" 
%                                                                      
%                                                           "with SEs" 
%                                                                      
%                                                  "0.501721750806993" 
%                                                                      
%                                                        "and 95 % CI" 
%                                                                 2.5% 
%                                                  "-4.38380704259523" 
%                                                                97.5% 
%                                                  "-2.40372605889863"
